\vspace{-1mm}
\section{Conclusions and Future Work\label{section-conclusions}}

We have demonstrated that Make really is unsuitable for large complex build
systems, regardless of whether used recursively or non-recursively. Using Shake
we have rewritten the GHC build system, producing the fifth and hopefully final
version. While all previous versions have started simple and gained complexity
as they progressed, this version is different. Developing the abstractions in
\S\ref{sec:abstractions} took many months of discussion and refinement. Once the
fundamental concepts were in place, the rest was ``just'' coding and reverse
engineering the existing system. The result is faster, more maintainable and
more correct. There are three major tasks remaining for future work:

\begin{itemize}
\item While we have demonstrated that our approach works, we have not yet
implemented all features of the build system, and hope to do so over the next
few months. Once complete, we expect it to quickly become the only supported
method of building GHC.

\item Our abstractions from \S\ref{sec:abstractions} were designed to allow
tracking provenance of command line arguments -- mapping each flag to the
location of the expression that generated it. This feature will rely on the
\emph{implicit locations} feature of the latest GHC.

% using the new \emph{implicit locations}
% feature\footnote{\scriptsize\urlstyle{sf}\url{https://ghc.haskell.org/trac/ghc/wiki/ExplicitCallStack/ImplicitLocations}$\!\!\!$}
% in GHC

\item While faster than the old system, the build is still slower than we would
like. The zero build time could be reduced by switching to a faster serialisation
library. The critical path of a full rebuild takes over seven minutes, limiting
the gains available from additional processors. We hope to break this critical path
by refactoring the build system, which is now a feasible task.
\end{itemize}
