\documentclass[preprint]{sigplanconf}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{balance}
\usepackage{verbatim}
\usepackage[T1]{fontenc}  % access \textquotedbl
\usepackage{textcomp}
\usepackage[all]{xy}
\include{paper}

\lstset{
    basicstyle=\sffamily,
    columns=fullflexible,
    keepspaces=true,
    escapechar=\#,
    literate={-}{{\hbox{-}}}1
             {_}{{\hbox{\tiny\_}}}1
             {`}{{$^\backprime$}}1
             {"}{{\textquotedbl}}1
             {->}{{$\rightarrow$}}1 {<-}{{$\leftarrow$}}1
             {=>}{{$\Rightarrow$}}1
             {>=}{{$\geq$}}1
             {<=}{{$\leq$}}1
             {>>}{{>\!>}}1
             {<<}{{<\!<}}1
             {<>}{{<\hspace{-0.4mm}>}}1
             {</>}{{<\hspace{-0.4mm}/\hspace{-0.4mm}>}}1
             {<.>}{{<\hspace{-0.3mm}.\hspace{-0.3mm}>}}1
             {<\$>}{{<\hspace{-0.3mm}\$\hspace{-0.3mm}>}}1
             {<*>}{{<\hspace{-0.3mm}*\hspace{-0.3mm}>}}1
             {&&&}{{$\wedge$}}1
             {|||}{{$\vee$}}1
             {!!!}{{$\neg$}}1
             {===}{{$\equiv$}}1 }
\newcommand\lst\lstinline

\makeatletter
\def \@ivtitleauthors#1#2#3#4{%
  \if \@andp{\@emptyargp{#2}}{\@emptyargp{#3}}%
    \noindent \@setauthor{40pc}{#1}{\@false}\par
  \else\if \@emptyargp{#3}%
    \noindent \@setauthor{17pc}{#1}{\@false}\hspace{3pc}%
              \@setauthor{17pc}{#2}{\@false}\par
  \else\if \@emptyargp{#4}%
    \noindent \@setauthor{17pc}{#1}{\@false}\hspace{3pc}%
              \@setauthor{17pc}{#3}{\@false}\par
  \else
    \noindent \@setauthor{9.3333pc}{#1}{\@false}\hspace{1.5pc}%
              \@setauthor{9.3333pc}{#2}{\@false}\hspace{1.5pc}%
              \@setauthor{9.3333pc}{#3}{\@false}\hspace{1.5pc}%
              \@setauthor{9.3333pc}{#4}{\@true}\par
    \relax
  \fi\fi\fi
  \vspace{20pt}}
\def \@maketitle {%
  \begin{center}
  \@settitlebanner
  \let \thanks = \titlenote
  {\leftskip = 0pt plus 0.25\linewidth
   \rightskip = 0pt plus 0.25 \linewidth
   \parfillskip = 0pt
   \spaceskip = .7em
   \noindent \LARGE \bfseries \@titletext \par}
  \vskip 6pt
  \noindent \Large \@subtitletext \par
  \vskip 12pt
  \ifcase \@authorcount
    \@latex@error{No authors were specified for this paper}{}\or
    \@titleauthors{i}{}{}\or
    \@titleauthors{i}{ii}{}\or
    \@titleauthors{i}{ii}{iii}\or
    \@ivtitleauthors{i}{ii}{iii}{iv}\or
    \@titleauthors{i}{ii}{iii}\@titleauthors{iv}{v}{}\or
    \@titleauthors{i}{ii}{iii}\@titleauthors{iv}{v}{vi}\or
    \@titleauthors{i}{ii}{iii}\@titleauthors{iv}{v}{vi}%
                  \@titleauthors{vii}{}{}\or
    \@titleauthors{i}{ii}{iii}\@titleauthors{iv}{v}{vi}%
                  \@titleauthors{vii}{viii}{}\or
    \@titleauthors{i}{ii}{iii}\@titleauthors{iv}{v}{vi}%
                  \@titleauthors{vii}{viii}{ix}\or
    \@titleauthors{i}{ii}{iii}\@titleauthors{iv}{v}{vi}%
                  \@titleauthors{vii}{viii}{ix}\@titleauthors{x}{}{}\or
    \@titleauthors{i}{ii}{iii}\@titleauthors{iv}{v}{vi}%
                  \@titleauthors{vii}{viii}{ix}\@titleauthors{x}{xi}{}\or
    \@titleauthors{i}{ii}{iii}\@titleauthors{iv}{v}{vi}%
                  \@titleauthors{vii}{viii}{ix}\@titleauthors{x}{xi}{xii}%
  \else
    \@latex@error{Cannot handle more than 12 authors}{}%
  \fi
  %\vspace{1pc}
  \end{center}}
\makeatother

\begin{document}

\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{Haskell Symposium 2016}{September 22--23, 2016, Nara, Japan}
\copyrightyear{2016}
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
%\copyrightdoi{nnnnnnn.nnnnnnn}
% Uncomment the publication rights you want to use.
%\publicationrights{transferred}
%\publicationrights{licensed}     % this is the default
%\publicationrights{author-pays}
%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Non-recursive Make Considered Harmful}
\subtitle{Build Systems at Scale}

\authorinfo{Andrey Mokhov\thanks{Andrey Mokhov conducted this work during a 6-month research visit to Microsoft Research Cambridge which was funded by Newcastle University, EPSRC (grant reference EP/K503885/1), and Microsoft Research.}}{Newcastle University}{andrey.mokhov@ncl.ac.uk}
\authorinfo{Neil Mitchell\thanks{Neil Mitchell is employed by Standard Chartered Bank. This paper has been created in a personal capacity and Standard Chartered Bank does not accept liability for its content. Views expressed in this paper do not necessarily represent the views of Standard Chartered Bank.}}{Standard Chartered Bank}{ndmitchell@gmail.com}
\authorinfo{Simon Peyton Jones}{Microsoft Research}{simonpj@microsoft.com}
\authorinfo{Simon Marlow}{Facebook}{smarlow@fb.com}

\maketitle
\begin{abstract}
Most build systems start small and simple, but over time grow into hairy monsters
that few dare to touch. As we demonstrate in this paper, there are a few issues
that cause build systems major scalability challenges, and many pervasively
used build systems (e.g. Make) do not scale well.

This paper presents a solution to the challenges we identify. We use functional
programming to design abstractions for build systems, and implement
them on top of the Shake library, which allows us to describe build rules and
dependencies. To substantiate our claims, we engineer a new build system
for the Glasgow Haskell Compiler. The result is more scalable, faster, and
spectacularly more maintainable than its Make-based predecessor.
\end{abstract}

%\category{CR-number}{subcategory}{third-level}
%\keywords
%build-system, compilation, Haskell

\input{introduction.tex}
\input{problems.tex}
\input{shake.tex}
\input{solutions.tex}
\input{abstractions.tex}
\input{GHC.tex}
\input{review.tex}
\input{conclusions.tex}

% \acks
% Thanks to Standard Chartered, where Shake was initially developed.
% \vspace{2mm}
% \noindent

\bibliographystyle{plainnat}
\balance
\bibliography

\end{document}
